Icicle: a single-pass query language

### 2015-12-16 2016icicle/section/02-Source.tex
running example and work on explicitly clocked syntax

Example data
\begin{code}
CODE, OPEN, CLOSE, DATE
ABC,    10,    12, 2015-01-01
ABC,    12,    14, 2015-01-02
IAG,    10,    12, 2015-01-01
IAG,    12,    14, 2015-01-02
\end{code}

Icicle queries execute over a single table, which must at least contain an entity identifier and a date.
Each query implicitly groups the data by entity and sorts it by date, before iterating through the data and computing some aggregate.
We define the stocks table as follows, leaving out the company name (entity identifier) and date.

@@

\begin{tabbing}
MM \= MM \= MMMMMMMMMMM \= \kill
$@definitions@$ \\
$\{$
\end{tabbing}

@@

We denote clocks in brackets as they are a different universe to normal expressions.
\begin{tabbing}
MM \= MM \= MMMMMMMMMMM \= \kill
$\mi{sum}[c]~v$    \\
\> $=$  \> $@fold@[c]~s~=~0~@then@~s~+~v$ \\
\> @in@ \> $s$. \\
\end{tabbing}

@@

Next, we can define $\mi{count}$ and $\mi{mean}$ in terms of this sum.
For 
\begin{tabbing}
MM \= MM \= MMMMMMMMMMM \= \kill
$\mi{count}$                                        \\
 \> $=$  \> $\mi{sum}~1$.                           \\
                                                    \\
$\mi{mean}~v$                                       \\
 \> $=$  \> $\mi{sum}~v~/~\mi{count}$.              \\
\end{tabbing}

We can use @filter@ to count only the days where the close price is higher than the open.
\begin{tabbing}
MM \= MM \= MMMMMMMMMMM \= \kill
$\mi{gap}$                                          \\
 \> $=$  \> $@filter@~\mi{close}~>~\mi{open}$       \\
 \> @in@ \> $\mi{count}$                            \\
                                                    \\
$\mi{proportion}$                                   \\
 \> $=$  \> $\mi{gap}~/~\mi{count}$
\end{tabbing}

Finally, we can use @group@ to put into buckets based on the percentage of growth.
Here the key for the map is the percentage of growth, and the value is the number of days with that amount of growth, divided by the total number of days.
\begin{tabbing}
MM \= MM \= MMMMMMMMMMM \= \kill
$\mi{growthPercent}$                                        \\
 \> $=$  \> $((\mi{close}-\mi{open}) \times 100 ) / \mi{open}$  \\
                                                            \\
$\mi{growthBy}$                                             \\
 \> $=$  \> $@let@~\mi{total} = \mi{count}$                 \\
 \> @in@ \> $@group@~\mi{growthPercent}$                    \\
 \> @in@ \> $\mi{count} / \mi{total}$                    \\
\end{tabbing}

@@

This language is not referentially transparent, which may seem odd, but this actually allows us to restrict the language and uphold our performance guarantees.
The input stream is implicitly passed as an argument to all parts of the program, but cannot be referred to explicitly.
Additionally, operations like @filter@ create a new context where the input stream is modified.
This means that a @let@ binding outside of a @filter@ can have different semantics when moved inside the @filter@, despite having the same body.
However, by passing this argument implicitly we remove a class of programs that would be hard to guarantee performance for.
By prohibiting the programmer from accessing the underlying stream, they cannot perform certain operations like zipping two streams with different filters together.

>>>>

Icicle queries execute over a single table, which must at least contain an entity identifier and a date.
Each query implicitly groups the data by entity and sorts it by date, before iterating through the data and computing some aggregate.
We define the stocks table as follows, leaving out the company code (entity identifier) and date.

@@

@@

We denote clocks in brackets as they are a different universe to normal expressions.
\begin{tabbing}
MM \= MM \= MMMMMMMMMMM \= \kill
@function@ 
$\mi{sum}[c]~v$    \\
\> $=$  \> $@fold@[c]~s~=~0~@then@~s~+~v$ \\
\> @in@ \> $s$. \\
\end{tabbing}

@@

We can define a query that computes the sum of each company's open prices.
Query definitions are similar to function definitions, except they only take one argument: the clock describing all available input.
They also must return aggregates, as an element query could be unbounded in size.
\begin{tabbing}
MM \= MM \= MMMMMMMMMMM \= \kill
@query@ 
$\mi{opensum}[c]$    \\
\> $=$  \> $\mi{sum}[c]~\mi{open}$. \\
\end{tabbing}

We can run $\mi{opensum}$ over some example data.
This example data has the company code and date for each entry, as well as the stock's open and close prices.
Note that the sum only becomes available at the end of each entity, and only applies to a particular entity's elements.

\begin{code}
CODE, OPEN, CLOSE, DATE,        OPENSUM
ABC,    10,    12, 2015-01-01,       10
IAG,    11,    12, 2015-01-01
IAG,    12,    10, 2015-01-02,       23
\end{code}

Next, we can define functions $\mi{count}$ and $\mi{mean}$ in terms of $\mi{sum}$.
Count passes the argument $1$ to sum, and this is interpreted as an element containing all ones.
The input clocks are passed along as-is.
\begin{tabbing}
MM \= MM \= MMMMMMMMMMM \= \kill
@function@ 
$\mi{count}[c]$                                     \\
 \> $=$  \> $\mi{sum}[c]~1$.                           \\
                                                    \\
@function@ 
$\mi{mean}[c]~v$                                       \\
 \> $=$  \> $\mi{sum}[c]~v~/~\mi{count}[c]$.              \\
\end{tabbing}

Clocks can be reduced to filter out certain elements using @where@.
In this query, we use @where@ to count the number of days where the close price is higher than the open price.
\begin{tabbing}
MM \= MM \= MMMMMMMMMMM \= \kill
@query@ 
$\mi{gap}[c]$                                        \\
 \> $=$  \> $\mi{count}[c~@where@~\mi{close}~>~\mi{open}]$.       \\
\end{tabbing}

In the example below we have shown the result of the element expression $\mi{close}~>~\mi{open}$ for clarity only; this is not included in the output of the $\mi{gap}$ query.
\begin{code}
CODE, OPEN, CLOSE, CLOSE > OPEN, GAP
ABC,    10,    12,            T,   1
IAG,    11,    12,            T
IAG,    12,    10,            F,   1
\end{code}

@@
